---
title: "Visualize Data"
output:
  html_document:
    toc: true
    df_print: paged
---

<!-- This file by Shannon Dunnigan was adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->

"r setup" is an optional label for a chunk, but it should always run once before anything else. In this setup we are loading the `tidyverse` and `here` packages. 

```{r setup}
library(tidyverse)
library(here)
```

The `wq` data frame contains monthly averages of water quality data (e.g., temperature, salinity, dissolved oxygen, etc.) from six National Estuarine Research Reserves in the contiguous United States. *Notice the use of the `here` package in the code below.*

```{r loaddata}
wq <- read.csv(here::here('data', 'monthly_wq.csv')) 

```

**Tip**: Always be sure to check your data. You can use `head()`, `tail()`, or `str()` from the base package in `R`. Or one of our favorite quick an easy ways is `dplyr::glimpse()`. The `glimpse()` function from the `dplyr` package is a great all in one snapshot of the data.


### Your Turn 1

Run the code on the slide to make a graph. Pay strict attention to spelling, capitalization, and parentheses!

```{r}


```

### Your Turn 2a

Add `color`, `size`, `alpha`, and `shape` aesthetics to your graph. Experiment.  

```{r}
# original
ggplot(data = wq) + 
  geom_point(mapping = aes(x = temp, y = do_mgl))


```

### Your Turn 2b

Using the `wq` data, plot temperature (`temp`) by `month` as a scatterplot and facet by `state` and `coast`. 
Play around with `facet_grid()` and `facet_wrap()`. 

*What kinds of things do you notice?*

```{r}


```


### Your Turn 3 - Zoom whiteboard responses

Replace this scatterplot with one that draws boxplots. Use the cheatsheet. Try your best guess.

```{r}
ggplot(data = wq) +
  geom_point(mapping = aes(x = coast, y = temp)) 

```


### Your Turn 4

Load the `diamonds` dataset and examine it. *ahem* Remember the `glimpse()` function? This data set contains prices and other attributes of diamonds and is provided by the `ggplot2` package.

Create a bar chart of `cut` colored by `cut`. Use the cheatsheet. 

```{r}
diamonds

```

### Your Turn 5 - Zoom whiteboard

Predict what this code will do. Then run it.

```{r}
ggplot(wq) +
  geom_point(aes(temp, do_mgl)) +
  geom_smooth(aes(temp, do_mgl)) 

```

# To Recap:

To make a graph:

1.    Pick your *data* set
2.    Choose the best *geom* to display the cases
3.    *Map* aesthetic properties to variables.
4.    **Not required**, but defaults are supplied by the specified geom that you can then manually change.

```{r, eval = FALSE}
ggplot(data = <DATA>) +
	<GEOM FUNCTION>(mapping = aes(<MAPPINGS>),
				   stat = <STAT>, # not required, but sensible default applied.
				   position = <POSITION>) # not required, but sensible default applied.
```

## Saving your plots

*What does `getwd()` return?* Zoom whiteboard

```{r}
getwd()
```


## DAY 1 CHALLENGE

Using the `diamonds` data make a scatterplot of `carat` by `price` and assign color by `clarity`. 
Customize the themes and colors of the plot to make it the *ugliest* plot that you can. 
Save this plot in your files pane.

``` {r ugly}
ggplot(data = diamonds) +
  geom_point(aes(x = carat, y = price, color = clarity))
```

***

# Take aways

You can use this code template to make thousands of graphs with `ggplot2`.

```{r, eval = FALSE}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>),
                  stat = <STAT>,
                  position = <POSITION>) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION> +
  <SCALE_FUNCTION> +
  <THEME_FUNCTION>
```

*** 

# For the sake of time

Here are some areas for more practice on your own:

### Find the default stat for geoms

Determine what the default stat is for each of the following geoms (you can use the cheatsheet OR the help function in RStudio):

-  `geom_line`
-  `geom_density`
-  `geom_smooth`

```{r, eval = FALSE}
?geom_line
?geom_density
?geom_smooth
```

## Saving plots

Save the last plot and then locate it in the files pane.

```{r, eval = FALSE}

```