---
title: "Functions"
author: "Cressman and Dunningan | Integrate R | June 2020"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
    df_print: tibble
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Setup:

```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(tidyr)
library(here)
library(ggplot2)

ebird <- read.csv(here::here("data", "eBird_workshop.csv"), stringsAsFactors = FALSE)
ebird <- dplyr::distinct(ebird)

wq <- read.csv(here::here("data", "daily_wq.csv"), stringsAsFactors = FALSE)
wq_trimmed <- wq %>%
    select(station_code, month, day, temp, sal, do_pct, depth) %>%
    filter(!is.na(depth)) %>%
    mutate(depth_ft = depth * 3.28,
           temp_f = (9/5) * temp + 32)
```


# Functions  

**DRY** - **D**on't **R**epeat **Y**ourself.  

If you've used the same chunk of code more than 3 times, you may want to consider writing a *function*. This can reduce errors due to typos (or just forgetting to change something in some spot), and also makes it easier to re-run a whole pile of code when you realize something needed to change - you only have to change it in one place!    


## Function Example  

R has functions to calculate mean and standard deviation of a dataset, but there isn't one for standard error. Remember, `standard error = sd / sqrt(n)`. 

Let's get this working on a small chunk of data. I like to pull out a little bit and call it "test".   

```{r}
test <- wq_trimmed %>% 
  filter(station_code == "gndblwq",
         month == 1)    
```

Then I get the code working on my test data:

```{r}
sd(test$temp_f, na.rm = TRUE) / sqrt( sum(!is.na(test$temp_f)) )
```


See why that might be annoying to type over and over again? And how it could be prone to error?  


The thing we're repeating, that we might want to generalize, is `test$temp_f`.  


So let's assign that to something, and replace the specific value in our code with the variable `x`:  

```{r}
x <- test$temp_f
sd(x, na.rm = TRUE) / sqrt( sum(!is.na(x)) )
```


We're most of the way to the function. We need to wrap it up like this, in `()` and `{}`, and come up with a name. RStudio can help with this formatting, if you go to the 'Code' toolbar and select 'Extract Function'.    

```{r}
sterr <- function(x){
  sd(x, na.rm = TRUE) / sqrt( sum(!is.na(x)) )
}
```


Run the above chunk of code. Now you can see `sterr` in your Environment pane as a function!  And you use it like any other function:  

```{r}
sterr(test$temp_f)
sterr(test$sal)
```


Let's check it on our bigger data frame!  

```{r}
sterr(wq_trimmed$temp_f)
```


# Your Turn 1  

## 1a  

Now **you** write a function! Name it `divide_by_10`. The input should be a vector (like we did with `sterr`), and the output should be each value of that vector divided by 10.  

Remember to get it working *before* you wrap it up into a function.  

```{r}







```


## 1b   

Now, what if we want to divide by something other than 10? Generalize the function to take two arguments as input. The second input (call it `y`) should be the denominator in your `divide_by` function.  


```{r}







```

Reminder: It's always a good idea to make sure you get output you expect!  




## Setting default values for arguments  

The next thing we'll do is set a default argument for `y` - if nothing else is specified, our function will divide the input vector by 10. This is done in the argument definition:  

```{r}
divide_by <- function(x, y = 10){
  x / y
}
```

And we can test it the same way we've tested others.  

```{r}
divide_by(test$sal, 10)
divide_by(test$sal)
divide_by(test$sal, 100)
```



***  
***  


# Functions using the tidyverse  

What we've done in this workshop has been very tidyverse-heavy. Functions using the tidyverse need to be handled a little differently. We will use graphing as an example.    

First we'll make a simple plot. I'm not claiming that this is pretty.  

```{r}
ggplot(wq_trimmed) +
  geom_point(aes(x = sal, 
                 y = do_pct, 
                 col = station_code))
```


Say we want to make this plot for a variety of pairs of parameters, with a third to provide color. So these are three arguments that we want to be able to specify. When we use RStudio's menu to extract a function, it gives us four - it includes the data frame as the first argument (it may not always be called `wq_trimmmed`) - and it looks like this:    

```{r}
my_plot <- function(wq_trimmed, sal, 
                    do_pct, station_code) {
  ggplot(wq_trimmed) +
    geom_point(aes(x = sal, 
                   y = do_pct, 
                   col = station_code))
}
```


So let's change some names and see what happens.  

```{r}
my_plot <- function(data, param1, 
                    param2, param3) {
  ggplot(data) +
    geom_point(aes(x = param1, 
                   y = param2, 
                   col = param3))
}
```


Try.....

```{r, eval = FALSE}
my_plot(wq_trimmed, sal, do_pct, station_code)
```


Uh-oh.  

Fortunately, the solution is simple: wrap the arguments that will change inside two curly braces `{{ }}`.  

```{r}
my_plot <- function(data, param1, 
                    param2, param3) {
  ggplot(data) +
    geom_point(aes(x = {{ param1 }}, 
                   y = {{ param2 }}, 
                   col = {{ param3 }})) 
}
```

And play with it a bit.  

```{r}
my_plot(wq_trimmed, sal, do_pct, station_code)
my_plot(wq_trimmed, sal, do_pct, month)  

my_plot(wq_trimmed, sal, do_pct, factor(month))  
# try it yourself!  
```


If in doubt, "embrace" it! `{{ }}`



# Your Turn 2  

Add theme elements to the plot in your `my_plot` function to make it a plot you actually *want* to reproduce. Then use your function on at least two combinations of parameters (we suggest `do_pct` vs `temp`, and `do_pct` vs `sal`).  

```{r}





```


